---
title: "DATA1001 Project"
author: "540539824, 550668899, 440318530, 541006846, 530657501"  
format: 
  html:
    embed-resources: true
    toc: true
    code-fold: true
    theme: sandstone
    highlight-style: github
    css: style.css  # custom css file
    
---

# Recommendation/Insight

According to our results, the rental prices in Sydney have increased over time regardless of where students live. We recommend that the university provide more affordable housing options to make education more accessible and affordable for students. 

# Evidence

## IDA

This dataset was sourced from a voluntary survey completed by students who were or currently are enrolled in DATA1X01 at the University of Sydney. 

Analysis was focused on three key variables: weekly rent paid in AUD (quantitative, continuous), commute time to campus in minutes (quantitative, continuous), and student cohort—2024S2 or 2025S1 (qualitative, discrete). 

The data only considered DATA1X01 students and neither reflected the whole student body, nor all students in Sydney. It was also self-reported and unverified, which may have made it inaccurate as respondents could have misinterpreted questions or lied. 

It was assumed that people honestly and reasonably reported the amount of rent they paid, and also an estimate of the time it takes to commute to campus. “$0” rent meant that the student did not pay rent, and it was assumed they lived with parents or family. These were excluded from the dataset as they don’t provide useful information about rent prices. Rents above $2000 were also excluded, as they could represent shared leases, data-entry errors, or luxury rentals not representing student populations. Commute times above 3 hours were also not considered, as 3 hours is a generous upper limit for reasonable commutes.

```{r, warning = FALSE, message = FALSE}
library(tidyverse)      # for box and linear model
library(ggthemes)       # for ggplot theme
surveydata = read.csv("data1001_survey_data_2025_S1.csv")     
surveydata = filter(surveydata, consent == "I consent to take part in the study") # remove students who did not consent to take part in the study
surveydata = surveydata[surveydata$rent <= 2000, ]    # remove data with rent greater than 2000
surveydata = surveydata[surveydata$rent != 0, ]       # remove data with 0 rent
surveydata = surveydata[surveydata$commute <= 180, ]  # remove data with commute time greater than 180 minutes
```




## Research Question 1

**How have rent prices changed between Semester 2 of last year (2024) and Semester 1 this year (2025) ?**


```{r}
ggplot(surveydata, aes(x = rent, y = cohort)) +
  geom_boxplot() +
  labs(x = "Rent",
       y = "Semester") +
  theme_solarized() +
  scale_fill_solarized()
```

Rent prices for 2025S1 students showed a rise compared to 2024S2, this was reflected by an increased median (550 vs. 500), mean (557.46 vs. 526.83), and IQR (366 vs. 328). The higher median suggests rent inflation, while the larger IQR indicates a broader range of prices, this could suggest that there are possible external factors like economic shifts and housing demand changes that affected rent prices. KPMG's 6-monthly property report supports this as they explore how unit “rent prices will gain by 4.6% in 2025 as units represent a more realistic housing route, given the ongoing affordability crisis” (inflation)

However, data limitations such as the significantly different sample sizes (597 in 2024S2 vs. 1506 in 2025S1) and the short timeframe of two semesters, prevents the analysis of long-term trends. Additionally, responses reporting $0 were excluded, assuming these students live with their parents, they do not align with our research question. 


## Research Question 2

:::: {.panel-tabset}
### Scatter Plot
```{r, warning = FALSE, message = FALSE}
ggplot(surveydata, aes(x = commute, y = rent)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Commute",
       y = "Rent") +
  theme_solarized() +
  scale_fill_solarized()
```

### Residual Plot
```{r, warning = FALSE, message = FALSE}
model = lm(rent ~ commute, data = surveydata)

# Create residual plot
# Create residual plot
ggplot(model, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "red") +
  labs(x = "Fitted Value",
        y = "Residual") +
  theme_solarized() +
  scale_fill_solarized()
```
::::






## Articles

[1]Welch, I. (2025, January 28). House prices to rise by 3.3%, units by 4.6% in 2025. KPMG. https://kpmg.com/au/en/home/media/press-releases/2025/01/house-and-unit-prices-to-rise-in-2025.html

# Acknowledgements

The Acknowledgment section includes a list of group meetings (date and time and attendance), the contribution of each group member, and all resources used (eg url of stack overflow, url of Ed post, date and details of drop-in session with tutor, record of ChatGPT session with prompt)
 


